\section{Hartree-Fock Self-Consistent-Field} \label{scf}
                                                                                
Hartree-Fock molecular orbital theory forms the cornerstone of
ab initio quantum chemistry.  Until the advances in the accuracy
of Kohn-Sham density functional theory in the 1990's, Hartree-Fock
theory was the method of choice for obtaining results for large
molecules without resorting to standard empirical or semiempirical
approaches.  Molecular properties obtained by Hatree-Fock theory are
generally at least qualitatively correct, although they can be
quantitatively poor in many instances.
                                                                                
\PSIthree\ solves the Hatree-Fock equations in a basis of Gaussian 
functions using an iterative, self-consistent-field (SCF) procedure.  The
final molecular orbitals are those which minimize the energy, 
subject to the electron configuration specified by the user (or
guessed by the program).  It is important to point out that the
SCF approach does not rigorously guarantee that the final orbitals
actually correspond to a minimum in orbital space; at convergence,
the only guarantee is that the gradient of the energy with respect 
to orbital rotations is zero: this could be a global minimum, a local
minimum, or a saddle point in orbital rotation space.  While this
is not usually an issue, it can be a problem sometimes for radicals,
diradicals, bond breaking, or unusual bonding situations.  The
\PSIstable\ module can be used to test for the stability of Hartree-Fock
wave functions.

Of course the efficiency of the iterative procedure depends on the choice of
initial guess.  The \PSIcscf\ module will attempt to use previously obtained
orbitals as a guess if they are available.  This can be particularly
advantageous when diffuse functions are present; in that case, it may
be easiest to run the computation with a smaller basis and project
those orbitals onto the larger basis by specifying the \keyword{--chkptmos}
option when running the \PSIinput\ program for the larger basis.  If
old MO's are not available, \PSIcscf\ uses a core Hamiltonian guess
by default.

Convergence, DIIS.

Keywords.



