\section{A \PSIthree\ Tutorial} \label{tutorial}

\subsection{Before Getting Started: A Warning about Scratch Files}
Generally, electronic structure programs like \PSIthree\ make
significant use of disk drives.  Therefore, it is very important
to ensure that PSI3 is writing its temporary files to a disk drive
phsyically attached to the computer running the computation.  If it
is not, it will significantly slow down the program and the network.
By default, PSI3 will write temporary files to \file{/tmp}, but you
will want to set up a default scratch path (as described in sections
\ref{scratchfiles} and \ref{psirc}) because the \file{/tmp} directory
is usually not large enough except for small test cases.

\subsection{Basic Input File Structure} 

PSI3 reads input from a text file, which can be prepared in any standard
text editor.  The default input file name is input.dat and the default
output file name is output.dat.  So that you can give your files meaningful
names, these defaults can be changed by specifying the input file name and
output file name on the the command line.  The syntax is:

{\tt psi3 input-name output-name}

PSI3 is a modular program, with each module performing specific tasks
and computations during your calculation.  Which modules are run for
a particular computation depends on the type of computation and the
particular keywords specified in the input file.  All keywords in PSI3
use the structure {\tt keyword = value}, where values may be strings,
booleans, integers or real numbers.  You can give keywords in the
input file for specific modules; however, in the first few examples,
we will place all our keywords in one section of our input file called
{\tt psi}.  Generally, every module you run during your computation will
read the keywords in {\tt psi}, so you can place all your keywords in
this section if you choose to do so.

\subsection{Running a basic SCF calculation}
In our first example, we will consider a Hartree-Fock SCF computation
for the water molecule using a cc-pVDZ basis set.  We will specify the
geometry of our water molecule using a standard z-matrix.

\begin{verbatim}
psi:(
 label = "cc-pVDZ SCF H2O"
 jobtype = sp
 wfn = scf
 reference = rhf
 basis = "cc-pVDZ"
 zmat = (
   o
   h 1 0.957
   h 1 0.957 2 104.5
  )
 )

\end{verbatim}

In each computation, you can specify the type of wavefunction (keyword
{\tt wfn}), the reference wavefunction (keyword {\tt reference}), and
the type of calculation you want to perform (keyword {\tt jobtype}).
In the example above, we used a restricted Hartree-Fock (RHF) reference
in an SCF computation of a single-point energy.  To change the level
of electron correlation, you change the type of wavefunction using the
keyword {\tt wfn}.  In the example above, to perform an
MP2 computation, simply set {\tt wfn = mp2}.

\subsection{Geometry Optimization and Vibrational Frequency Analysis}
The above example was a simple single-point energy computation.
To perform a different type of computation, change the keyword {\tt
jobtype}.  In the example below, we will set up
a CCSD geometry optimization.  To illustrate a more flexible z-matrix
input, we will now define variables for the bond length and bond angle
(in the {\tt zvars} section).

\begin{verbatim}
% 6-31G** H2O Test optimization calculation

psi: (
  label = "6-31G** SCF H2O"
  jobtype = opt
  wfn = ccsd
  reference = rhf
  dertype = first
  basis = "6-31G**"
  zmat = (
    o
    h 1 roh
    h 1 roh 2 ahoh
  )
  zvars = (
    roh     0.96031231
    ahoh  104.09437511
  )
)
\end{verbatim}

Once you have optimized the geometry of a molecule, you might wish to
perform a frequency analysis to determine the nature of the stationary
point.  To do this, change the value of {\tt jobtype} to {\tt
freq}.  For an SCF frequeny calculation, you would also set {\tt
dertype = second} to compute the second derivatives analytically.
Unfortunately, analytical second derivitives are not available 
in \PSIthree\ for wavefunctions beyond SCF,
so instead use the highest order analytical derivitives
that are available for the type of wavefunction you have chosen.  This
information is given in Table \ref{table:methods}.  For our CCSD example, the
highest-order derivitives available are first, so {\tt dertype = first}.

\begin{verbatim}
% 6-31G** H2O Test computation of frequencies

psi: (
  label = "6-31G** SCF H2O"
  jobtype = freq
  wfn = ccsd
  reference = rhf
  dertype = first
  basis = "6-31G**"
  zmat = (
    o
    h 1 roh
    h 1 roh 2 ahoh
  )
  zvars = (
    roh     0.96031231
    ahoh  104.09437511
  )
)
\end{verbatim}

\subsection{More Advanced Input Options}
If you wish to add comments to your input file, you can start any line
with \% and the line will be a comment line.  This can make the input
file easier to understand because you can provide explainations about
each keyword.  Another way to make the input file more organized is
to seperate it into sections that correspond to particular modules
the calculation will use.  This can be particularly helpful for more
complicated calculations which utilize a lot of keywords.  In the example
below, a CCSD(T) calculation for the BH molecule is performed using a
cc-pVDZ basis set.  The keywords are divided into sections and several
new keywords are introduced, including ones to specify symmetry and
orbital occupations.  All keywords which specify orbitial occupations
are a list of integers enclosed in parenthesis.  These integers give the
number of orbitials which belong to each irreducible representation in
the point group.  The ordering of the irreps are those given by Cotton
in {\em Chemical Applications of Group Theory}.  In this example,
comment lines will be included to explain the new keywords used.

\begin{verbatim}
psi: (
  wfn = ccsd_t
  reference = rhf
)

default: (
  label = "BH cc-pVDZ CCSD(T)"

% Allocating memory for the calculation
  memory = (600.0 MB)

% charge and multp default to values of 0 and 1, respectively
  charge = 0
  multp = 1

% The program will generally guess the symmetry of the molecule, but
% it can be overridden.  Here we specify c2v because only d2h and its
% subgroups can be used by the program.
  symmetry = c2v

% Number of orbitals per irrep can be specified manually, if desired
  docc = (3 0 0 0)

% Freeze the 1A1 orbital (B 1s-like) in the CCSD(T) computation
  frozen_docc = (1 0 0 0)
)

% The input section contains information about the molecule and the basis
% set.  The geometry here is specified by cartesian coordinates, therefore
% the keyword units must also be included.
input: (
  basis = "cc-pVDZ"
  units = angstroms
  geometry = (
    ( b      0.0000        0.0000        0.0000)
    ( h      0.0000        0.0000        0.8000)
      )
  origin = (0.0 0.0 0.0)
)
% The modular input structure lets you specify convergence criteria for
% each part of the calcultion.  
scf: (
  maxiter = 100
  convergence = 11
)
\end{verbatim}

The final example of this tutorial demonstrates an example of a CAS-SCF
calculation.  CAS calculations require specification of several additional
keywords because you must specify which orbitals you wish to be in the
active space.  The notation and ordering for specifying CAS orbitals is the
same as for occupied orbitals.

\begin{verbatim}

% 6-31G** H2O Test CASSCF Energy Point

psi: (
  label = "6-31G** CASSCF H2O"
  jobtype = sp
  wfn = casscf
  reference = rhf
% The restricted_docc orbitals are those which are optimized, but are not
% in the active space.
  restricted_docc = (1 0 0 0)

% The active space orbitals; this combination of restricted and active
% orbitals makes a valence active space for the water molecule.
  active          = (3 0 1 2)

  basis = "6-31G**"
  zmat = (
    o
    h 1 1.00
    h 1 1.00 2 103.1
  )
)
\end{verbatim}

