% PSI3 Programmer's Manual
%
% Binary I/O --- libpsio
%
% T. Daniel Crawford, June 2000
%

\subsubsection{The structure and philosophy of the library}

Almost all \PSIthree\ modules must exchange data with raw binary (also
called ``direct-access'') files.  However, rather than using low-level
C or Fortran functions such as \celem{read()} or \celem{write()},
\PSIthree\ uses a flexible, but fast I/O system that gives the
programmer and user control over the organization and storage of data.
Some of the features of the PSI I/O system, libpsio, include:
\begin{itemize}
\item A user-defined disk striping system in which a single binary
file may be split across several physical or logical disks.
\item A file-specific table of contents (TOC) which contains
file-global starting and ending addresses for each data item.
\item An entry-relative page/offset addressing scheme which avoids
file-global file pointers which can limit file sizes.
\end{itemize}

The TOC structure of PSI binary files provdes several advantages over
older I/O systems.  For example, data items in the TOC are identified
by keyword strings (e.g., \celem{"Nuclear Repulsion Energy"}) and the
{\em global} address of an entry is known only to the TOC itself,
never to the programmer. Hence, if the programmer wishes to read or
write an entire TOC entry, he/she is required to provide only the TOC
keyword and the entry size (in bytes) to obtain the data.
Furthermore, the TOC makes it possible to read only pieces of TOC
entries (say a single buffer of a large list of two-electron
integrals) by providing the appropriate TOC keyword, a size, and a
starting address relative to the beginning of the TOC entry. In short,
the TOC design hides all information about the global structure of the
direct access file from the programmer and allows him/her to be
concerned only with the structure of individual entries. The current
TOC is written to the end of the file when it is closed.

Thus the direct-access file itself is viewed as a series of pages,
each of which contains an identical number of bytes. The global
address of the beginning of a given entry is stored on the TOC as a
page/offset pair comprised of the starting page and byte-offset on
that page where the data reside. The entry-relative page/offset
addresses which the programmer must provide work in exactly the same
manner, but the 0/0 position is taken to be the beginning of the TOC
entry rather than the beginning of the file.

\subsubsection{The user interface}
All of the functions needed to carry out basic I/O are described in
this subsection. Proper declarations of these routines are provided by
the header file \file{psio.h}. Note that before any open/close
functions may be called, the input parsing library, libipv1 must be
initialized so that the necessary file striping information may be
read from user input, but this is hidden from the programmer in
lower-level functions.  NB, \celem{ULI} is used as an abbreviation for
\celem{unsigned long int} in the remainder of this manual.

\celem{int psio\_init(void)}: Before any files may be opened or the
basic read/write functions of libpsio may be used, the global data
needed by the library functions must be initialized using this
function.

\celem{int psio\_ipv1\_config(void)}: For the library to operate properly,
its configuration must be read from the input file or from user's {\tt .psirc} file.
This call MUST immediately follow {\em int psio\_init();}.

\celem{int psio\_done(void)}: When all interaction with the
direct-access files is complete, this function is used to free the
library's global memory.

\celem{int psio\_open(ULI unit, int status)}: Opens the direct access
file identified by \celem{unit}. The \celem{status} flag is a boolean
used to indicate if the file is new (0) or if it already exists and is
being re-opened (1). If specified in the user input file, the file
will be automatically opened as a multivolume (striped) file, and each
page of data will be read from or written to each volume in
succession.

\celem{int psio\_close(ULI unit, int keep)}: Closes a direct access
file identified by unit. The keep flag is a boolean used to indicate
if the file's volumes should be deleted (0) or retained (1) after
being closed.

\celem{int psio\_read\_entry(ULI unit, char *key, char *buffer, ULI
size)}: Used to read an entire TOC entry identified by the string
\celem{key} from \celem{unit} into the array \celem{buffer}. The
number of bytes to be read is given by \celem{size}, but this value is
only used to ensure that the read request does not exceed the end of
the entry. If the entry does not exist, an error is printed to stderr
and the program will exit.

\celem{int psio\_write\_entry(ULI unit, char *key, char *buffer, ULI
size)}: Used to write an entire TOC entry idenitified by the string
\celem{key} to \celem{unit} into the array \celem{buffer}. The number
of bytes to be written is given by \celem{size}. If the entry already
exists and its data is being overwritten, the value of size is used to
ensure that the write request does not exceed the end of the entry.

\celem{int psio\_read(ULI unit, char *key, char *buffer, ULI size,
psio\_address sadd, psio\_address *eadd)}: Used to read a fragment of
\celem{size} bytes of a given TOC entry identified by \celem{key} from
\celem{unit} into the array \celem{buffer}. The starting address is
given by the \celem{sadd} and the ending address (that is, the
entry-relative address of the next byte in the file) is returned in
\celem{*eadd}.

\celem{int psio\_write(ULI unit, char *key, char *buffer, ULI size,
psio\_address sadd, psio\_address *eadd)}: Used to write a fragment of
\celem{size} bytes of a given TOC entry identified by \celem{key} to
\celem{unit} into the array \celem{buffer}. The starting address is
given by the \celem{sadd} and the ending address (that is, the
entry-relative address of the next byte in the file) is returned in
\celem{*eadd}.

The page/offset address pairs required by the preceeding read and
write functions are supplied via variables of the data type
\celem{psio\_address}, defined by:
\begin{verbatim}
  typedef struct {
    ULI page;
    ULI offset;
  } psio_address;
\end{verbatim}
The \celem{PSIO\_ZERO} defined in a macro provides a convenient input
for the 0/0 page/offset.

\subsubsection{Manipulating the table of contents}
In addition, to the basic open/close/read/write functions described above,
the programmer also has a limited ability to directly manipulate or examine
the data in the TOC itself.

\celem{int psio\_tocprint(ULI unit, FILE *outfile)}: Prints the TOC of
\celem{unit} in a readable form to \celem{outfile}, including entry
keywords and global starting/ending addresses.  (\celem{tocprint} is
also the name of a \PSIthree\ utility module which prints a file's TOC to
stdout.)

\celem{int psio\_toclen(ULI unit, FILE *outfile)}: Returns the number
of entries in the TOC of \celem{unit}.

\celem{int psio\_tocdel(ULI unit, char *key)}: Deletes the TOC entry
corresponding to \celem{key}. NB that this function only deletes the
entry's reference from the TOC itself and does not remove the
corresponding data from the file. Hence, it is possible to introduce
data "holes" into the file.

\celem{int psio\_tocclean(ULI unit, char *key)}: Deletes the TOC entry
corresponding to \celem{key} and all subsequent entries. As with
\celem{psio\_tocdel()}, this function only deletes the entry
references from the TOC itself and does not remove the corresponding
data from the file. This function is still under construction.

\subsubsection{Using \library{libpsio.a}}
The following code illustrates the basic use of the library, as well
as when/how the \celem{psio\_init()}, \celem{psio\_ipv1\_config()}, and \celem{psio\_done()}
functions should be called in relation to initialization of
\library{libipv1}.  (See section \ref{C_Program} later in the manual
for a description of the basic elements of C-language \PSIthree\
program.)
\begin{verbatim}
#include <stdio.h>
#include <libipv1/ip_lib.h>
#include <libpsio/psio.h>
#include <libciomr/libciomr.h>

FILE *infile, *outfile;

int main()
{
  int i, M, N;
  double enuc, *some_data;
  psio_address next;  /* Special page/offset structure */

  ffile(&infile,"input.dat",2);
  ffile(&outfile,"output.dat",1);
  ip_set_uppercase(1);
  ip_initialize(infile,outfile);
  ip_cwk_add(":DEFAULT");
  ip_cwk_add(progid);

  /* Initialize the I/O system */
  psio_init(); psio_ipv1_config();

  /* Open the file and write an energy */
  psio_open(31, PSIO_OPEN_NEW);
  enuc = 12.3456789;
  psio_write_entry(31, "Nuclear Repulsion Energy", (char *) &enuc,
                   sizeof(double));
  psio_close(31,1);

  /* Read M rows of an MxN matrix from a file */
  some_data = init_matrix(M,N);

  psio_open(91, PSIO_OPEN_OLD);
  next = PSIO_ZERO;/* Note use of the special macro */
  for(i=0; i < M; i++)
      psio_read(91, "Some Coefficients", (char *) (some_data + i*N),
                N*sizeof(double), next, &next);
  psio_close(91,0);

  /* Close the I/O system */
  psio_done();

  ip_done();
}

char *gprgid()
{
   char *prgid = "CODE_NAME";
   return(prgid);
}
\end{verbatim}

The interface to the \PSIthree\ I/O system has been designed to mimic
that of the old \celem{wreadw()} and \celem{wwritw()} routines of
\library{libciomr} (see the next section of this manual).  The table
of contents system introduces a few complications that users of the
library should be aware of:
\begin{itemize}
\item As pointed out earlier, deletion of TOC entries is allowed using
\celem{psio\_tocdel()} and \celem{psio\_tocclean()}. However, since
only the TOC reference is removed from the file and the corresponding
data is not, a data hole will be left in the file if the deleted entry
was not the last one in the TOC. A utility function designed to
"defrag" a PSI file may become necessary if such holes ever present a
problem.
\item One may append data to an existing TOC entry by simply writing
beyond the entry's current boundary; the ending address data in the
TOC will be updated automatically. However, no safety measures have
been implemented to prevent one from overwriting data in a subsequent
entry thereby corrupting the TOC. This feature/bug remains because (1)
it is possible that such error checking functions may slow the I/O
codes significantly; (2) it may be occasionally desirable to overwrite
exiting data, regardless of its effect on the TOC. Eventually a
utility function which checks the validity of the TOC may be needed if
this becomes a problem, particularly for debugging purposes.
\end{itemize}


